\documentclass[12pt,a4paper]{article}

% Essential packages
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{fancyhdr}
\usepackage{titlesec}
\usepackage{tocloft}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{enumitem}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pgfgantt}
\usepackage{parskip}


% Page geometry
\geometry{
    top=2.5cm,
    bottom=2.5cm,
    left=2.5cm,
    right=2.5cm,
    headheight=15pt
}

% Color definitions
\definecolor{primaryblue}{RGB}{41, 128, 185}
\definecolor{secondaryblue}{RGB}{52, 152, 219}
\definecolor{darkgray}{RGB}{44, 62, 80}
\definecolor{lightgray}{RGB}{149, 165, 166}
\definecolor{accentgreen}{RGB}{39, 174, 96}
\definecolor{codebackground}{RGB}{248, 249, 250}

% Define team member colors
\definecolor{JesusGreen}{HTML}{27ae60}
\definecolor{ArnauPurple}{HTML}{9b59b6}
\definecolor{MarcBlue}{HTML}{2980b9}
\definecolor{MichelRed}{HTML}{e74c3c}
\definecolor{SharedOrange}{HTML}{FFA500}

% Hyperlink setup
\hypersetup{
    colorlinks=true,
    linkcolor=primaryblue,
    filecolor=primaryblue,
    urlcolor=primaryblue,
    citecolor=primaryblue,
    pdftitle={MD-P1-Report},
    pdfauthor={Group 6}
}

% Header and footer setup
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\textcolor{darkgray}{\small Data Mining}}
\fancyhead[R]{\textcolor{darkgray}{\small Group 6}}
\fancyfoot[C]{\textcolor{lightgray}{\thepage}}
\renewcommand{\headrulewidth}{0.5pt}
\renewcommand{\headrule}{\hbox to\headwidth{\color{primaryblue}\leaders\hrule height \headrulewidth\hfill}}

% Title page style
\fancypagestyle{titlepage}{
    \fancyhf{}
    \renewcommand{\headrulewidth}{0pt}
    \renewcommand{\footrulewidth}{0pt}
}

% Section title formatting
\titleformat{\section}
{\Large\bfseries\color{primaryblue}\raggedright}
{\thesection}
{1em}
{}
[\vspace{-0.5em}\textcolor{primaryblue}{\titlerule[1pt]}]

\titleformat{\subsection}
{\large\bfseries\color{secondaryblue}}
{\thesubsection}
{1em}
{}

\titleformat{\subsubsection}
{\normalsize\bfseries\color{darkgray}}
{\thesubsubsection}
{1em}
{}

% Table of contents styling
\renewcommand{\cftsecfont}{\color{primaryblue}\bfseries}
\renewcommand{\cftsubsecfont}{\color{secondaryblue}}
\renewcommand{\cftsecpagefont}{\color{primaryblue}\bfseries}
\renewcommand{\cftsubsecpagefont}{\color{secondaryblue}}

% Code listing setup
\lstdefinestyle{mystyle}{
    backgroundcolor=\color{codebackground},
    commentstyle=\color{lightgray},
    keywordstyle=\color{primaryblue},
    numberstyle=\tiny\color{lightgray},
    stringstyle=\color{accentgreen},
    basicstyle=\ttfamily\footnotesize,
    breakatwhitespace=false,
    breaklines=true,
    captionpos=b,
    keepspaces=true,
    numbers=left,
    numbersep=5pt,
    showspaces=false,
    showstringspaces=false,
    showtabs=false,
    tabsize=2,
    frame=single,
    frameround=tttt,
    rulecolor=\color{lightgray}
}
\lstset{style=mystyle}

% Custom title page commands
\title{}
\author{}
\date{}


\begin{document}

% Custom title page
\thispagestyle{titlepage}
\begin{center}

% Top spacing

% Main title with decorative elements
{\color{primaryblue}\rule{\textwidth}{2pt}}
\vspace{0.5cm}

{\Huge\bfseries\color{primaryblue} DATA MINING}
\vspace{0.3cm}

{\LARGE\color{secondaryblue} Food Marketing}
\vspace{0.5cm}

{\color{primaryblue}\rule{\textwidth}{2pt}}

\vspace{1.5cm}

% Project identifier
{\Large\bfseries\color{darkgray} P1-Report}

\vspace{2cm}

% Group information
{\large\bfseries\color{primaryblue} Group 6}

\vspace{1cm}

% Team members in a nice box
\fcolorbox{primaryblue}{white}{
\begin{minipage}{0.7\textwidth}
\centering
\vspace{0.5cm}
{\large\bfseries\color{primaryblue} Team Members}
\vspace{0.5cm}

\begin{tabular}{@{}c@{}}
\textcolor{darkgray}{Font I Cabarrocas, Marc} \\[0.3cm]
\textcolor{darkgray}{Garcia Ayala, Jesus} \\[0.3cm]
\textcolor{darkgray}{Hernández Navarro, Arnau} \\[0.3cm]
\textcolor{darkgray}{Mediavilla Jiménez, Alex Michel} \\
\end{tabular}
\vspace{0.5cm}
\end{minipage}
}

\vspace{1.5cm}

% Teacher information
{\large\bfseries\color{secondaryblue} Professor}
\vspace{0.3cm}

{\large\color{darkgray} Sergi Ramirez Mitjans}

\vspace{2cm}

% Date
{\large\color{lightgray} \today}

\vspace{1cm}

% Bottom decorative line
{\color{primaryblue}\rule{0.5\textwidth}{1pt}}

\end{center}
\newpage
\tableofcontents
\newpage

\section{Motivation}
Understanding consumer behaviour is crucial in the highly competitive food and grocery market. Consumers exhibit diverse preferences, purchasing habits, and responses to marketing efforts, influenced by demographic factors, lifestyle choices, and economic conditions. Analysing customer data allows businesses to move beyond generic strategies and tailor their offerings, promotions, and communication channels effectively. By uncovering patterns in how different customer groups interact with products (like wine, meat, fruit) and purchase channels (web, store, catalogue), we can gain valuable insights to enhance customer satisfaction, optimize marketing spend, and drive business growth. This project aims to dissect a rich dataset of customer interactions with a food retailer to reveal these underlying structures.


The primary problem addressed in this analysis is the identification and characterization of distinct customer segments within the provided food retail dataset (ifood\_enriched.csv).

While the raw data provides individual details, it's challenging to grasp the overarching patterns and group dynamics. Simply looking at averages or individual variables doesn't reveal how these attributes combine to form coherent customer profiles.

Therefore, the core task is to reduce the complexity of this multi-dimensional data and group similar customers together. We aim to answer questions like:

\begin{itemize}
    \item Are there natural groupings of customers based on their spending habits and demographics?
    \item What are the key characteristics that define these different customer segments?
    \item How do purchasing channel preferences differ across segments?
    \item Can we identify high-value vs. low-value segments, or perhaps niche groups with specific preferences?
\end{itemize}

The ultimate goal is to generate actionable insights by creating well-defined customer profiles based on the identified clusters. These profiles can then inform targeted marketing strategies, personalized recommendations, inventory management, and overall business decision-making for the food retailer

\newpage
\section{Data Source}
The dataset used in this project was obtained from Kaggle, an online platform for data science competitions and open datasets. Specifically, it comes from the dataset “Marketing Data” uploaded by Jack Daoud. The dataset contains customer information, including demographic details, purchasing behaviour, and responses to marketing campaigns, which makes it suitable for customer segmentation and predictive modelling.

% JESUS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\input{metadata.tex}


% ARNAU
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\input{descriptive.tex}

% MARC
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\input{pca.tex}

\newpage
\input{clustering.tex}

% MICHEL
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\input{profiling.tex}


% MARC
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\input{conclusions.tex}


% MARC
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\input{gant.tex}

\end{document}
