---
title: "Basic Descriptive Statistics — iFood CSV (DOCX‑safe)"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
  word_document: default
editor_options: 
  chunk_output_type: console
---

# Reference Descriptive Script (DOCX‑safe)

**Note:** Use forward slashes in paths on Windows (e.g., `C:/Users/...`).  
Place this `.Rmd` in the same folder as `ifood_base.csv` or update `data_path` below.

This version avoids HTML‑only features so it **knits to Word (.docx) without errors**.
If you knit to HTML, it will still show a floating ToC and nicer tables.

```{r setup, message=FALSE, warning=FALSE}
required_pkgs <- c("readr","dplyr","tidyr","stringr","purrr","lubridate","knitr")
to_install <- setdiff(required_pkgs, rownames(installed.packages()))
if (length(to_install)) install.packages(to_install, repos = "https://cloud.r-project.org")

library(readr); library(dplyr); library(tidyr); library(stringr); library(purrr)
library(lubridate); library(knitr)

# Safe Windows path
safe_path <- function(p) ifelse(is.na(p) || !nzchar(p), p, gsub("\\\\", "/", p))

# Table helper: uses HTML styling only when knitting to HTML; plain kable for Word/PDF
theme_table <- function(tbl) {
  if (knitr::is_html_output()) {
    if (!requireNamespace("kableExtra", quietly = TRUE)) {
      return(knitr::kable(tbl, align = "l"))
    }
    kableExtra::kbl(tbl, booktabs = TRUE, align = "l") |>
      kableExtra::kable_styling(full_width = FALSE, bootstrap_options = c("striped","hover","condensed"), font_size = 11)
  } else {
    knitr::kable(tbl, align = "l")
  }
}

# Plot colors (enough for most categorical vars)
listOfColors <- grDevices::rainbow(40)
```

## Read data

```{r read, message=FALSE, warning=FALSE}
# If your CSV is elsewhere, replace with an absolute path using forward slashes
# e.g., data_path <- "ifood_base.csv"
data_path <- "ifood_base.csv"
data_path <- safe_path(data_path)

if (!file.exists(data_path)) {
  stop(paste0("CSV not found at: '", data_path, "'. ",
              "If you're on Windows and used backslashes (e.g., C:\\Users\\...), ",
              "replace them with forward slashes like C:/Users/..."))
}

dd <- readr::read_csv(data_path, show_col_types = FALSE)
```

## Dimensions and variable names

```{r dims-names}
dim(dd)
n <- nrow(dd); K <- ncol(dd)
n; K

names(dd)
```

## Optional: Declare/standardize types (recommended for iFood)

```{r declare-types, message=FALSE, warning=FALSE}
# Normalize common alternative names
dd <- dd %>% rename(
  Year_Birth    = dplyr::any_of(c("Year_Birth","Year_birth","year_birth","Year","YearBirth")),
  Dt_Customer   = dplyr::any_of(c("Dt_Customer","DtCustomer","EnrollmentDate","Customer_Date","date_customer")),
  Marital_Status= dplyr::any_of(c("Marital_Status","Marital","marital_status")),
  Education     = dplyr::any_of(c("Education","education"))
)

# Parse date column if present
if ("Dt_Customer" %in% names(dd)) {
  dd$Dt_Customer <- suppressWarnings(parse_date_time(dd$Dt_Customer,
                            orders = c("dmy","ymd","mdy","d-b-Y","Y-m-d","d/m/Y","m/d/Y")))
  dd$Dt_Customer <- as.Date(dd$Dt_Customer)
}

# Categorical text columns
for (col in c("Education","Marital_Status")) {
  if (col %in% names(dd)) dd[[col]] <- as.factor(dd[[col]])
}

# Binary 0/1 columns (if present)
bin_cols <- intersect(c("AcceptedCmp1","AcceptedCmp2","AcceptedCmp3","AcceptedCmp4",
                        "AcceptedCmp5","Complain","Response"), names(dd))
for (bc in bin_cols) {
  if (all(na.omit(unique(dd[[bc]])) %in% c(0,1))) {
    dd[[bc]] <- factor(dd[[bc]], levels = c(0,1))
  }
}

str(dd)
```

## Missingness overview

```{r missingness}
miss_tbl <- tibble::tibble(
  variable = names(dd),
  missing_n = sapply(dd, function(x) sum(is.na(x))),
  missing_pct = round(100 * sapply(dd, function(x) mean(is.na(x))), 2)
) %>% arrange(desc(missing_pct))

theme_table(miss_tbl)
```

## Descriptive function

```{r descriptiva-fn}
descriptiva <- function(X, nom, nrows_total) {
  if (!(is.numeric(X) || inherits(X, "Date"))) {
    # Categorical
    frecs <- table(as.factor(X), useNA = "ifany")
    proportions <- frecs / nrows_total
    pie(frecs, cex = 0.7, main = paste("Pie of", nom))
    barplot(frecs, las = 3, cex.names = 0.7, main = paste("Barplot of", nom), col = listOfColors)
    cat("\nNumber of modalities:", length(frecs), "\n")
    cat("\nFrequency table\n"); print(frecs)
    cat("\nRelative frequency table (proportions)\n"); print(round(proportions, 4))
    cat("\nFrequency table sorted\n"); print(sort(frecs, decreasing = TRUE))
    cat("\nRelative frequency table (proportions) sorted\n"); print(round(sort(proportions, decreasing = TRUE), 4))
  } else {
    if (inherits(X, "Date")) {
      # Date
      cat("\nExtended Summary Statistics (Date)\n"); print(summary(X))
      sd_days <- tryCatch(sd(as.numeric(X), na.rm = TRUE), error = function(e) NA_real_)
      cat("\nsd (days):", round(sd_days, 3), "\n")
      hist(X, breaks = "weeks", main = paste("Histogram (weekly) of", nom), xlab = nom)
    } else {
      # Numeric
      hist(X, main = paste("Histogram of", nom), xlab = nom)
      boxplot(X, horizontal = TRUE, main = paste("Boxplot of", nom), xlab = nom)
      cat("\nExtended Summary Statistics\n"); print(summary(X))
      sd_x <- sd(X, na.rm = TRUE)
      mn_x <- mean(X, na.rm = TRUE)
      cat("\nsd:", round(sd_x, 6), "\n")
      cat("vc (sd/mean):", ifelse(is.na(mn_x) || mn_x == 0, NA, round(sd_x / mn_x, 6)), "\n")
    }
  }
}
```

## Run basic descriptives for all variables

```{r run-descriptive, fig.width=7, fig.height=4}
# Exclude known constants/IDs by default; adjust as needed
exclude <- intersect(c("Id","Z_CostContact","Z_Revenue"), names(dd))
actives <- setdiff(seq_len(ncol(dd)), match(exclude, names(dd)))

for (k in actives) {
  cat("\n\n## Variable", k, "—", names(dd)[k], "\n")
  descriptiva(dd[[k]], names(dd)[k], n)
}
```

## Appendix: Quick numeric summary table

```{r quick-numeric-summary, message=FALSE, warning=FALSE}
num_vars <- names(dd)[sapply(dd, is.numeric)]
if (length(num_vars)) {
  num_summary <- dd %>%
    summarise(across(all_of(num_vars),
                     list(min = ~min(.x, na.rm = TRUE),
                          q1  = ~quantile(.x, 0.25, na.rm = TRUE),
                          mean= ~mean(.x, na.rm = TRUE),
                          median= ~median(.x, na.rm = TRUE),
                          q3  = ~quantile(.x, 0.75, na.rm = TRUE),
                          max = ~max(.x, na.rm = TRUE),
                          sd  = ~sd(.x, na.rm = TRUE)), .names = "{.col}_{.fn}")) %>%
    pivot_longer(everything(), names_to = c("variable","stat"), names_sep = "_") %>%
    pivot_wider(names_from = stat, values_from = value)
  theme_table(num_summary)
} else {
  cat("No numeric variables detected.")
}
```