---
output:
  html_document: default
  pdf_document: default
  word_document: default
---

### *Load required libraries*
```{r}
# Suppress startup messages of library dplyr
suppressPackageStartupMessages(library(dplyr))
# Loading required libraries
library(dplyr, quietly = TRUE)
library(class, quietly = TRUE)
```
### 0. Load raw dataset
```{r}
ifood <- read.csv("ifood_base.csv", sep=",", header=TRUE, stringsAsFactors = FALSE)
```
### 1. Remove irrelevant columns
```{r}
ifood <- ifood[, !names(ifood) %in% c("ID", "Z_CostContact", "Z_Revenue")]
```
### 2. Transform date-related variables
```{r}
ifood$Age <- 2020 - ifood$Year_Birth
ifood <- ifood[, !names(ifood) %in% c("Year_Birth")]
reference_date <- as.Date("2020-12-31")
ifood$CustDays <- as.numeric(reference_date - as.Date(ifood$Dt_Customer, format="%Y-%m-%d"))
ifood <- ifood[, !names(ifood) %in% c("Dt_Customer")]
```
### 3. Rename columns for easier access
```{r}
colnames(ifood) <- gsub("NumDealsPurchases", "DealsPurc", colnames(ifood))
colnames(ifood) <- gsub("NumWebPurchases", "WebPurc", colnames(ifood))
colnames(ifood) <- gsub("NumStorePurchases", "StorePurc", colnames(ifood))
colnames(ifood) <- gsub("NumWebVisitsMonth", "WebVisits", colnames(ifood))
colnames(ifood) <- gsub("AcceptedCmpOverall", "CmpOverall", colnames(ifood))
colnames(ifood) <- gsub("MntWines", "WineExp", colnames(ifood))
colnames(ifood) <- gsub("MntFruits", "FruitExp", colnames(ifood))
colnames(ifood) <- gsub("MntMeatProducts", "MeatExp", colnames(ifood))
colnames(ifood) <- gsub("MntFishProducts", "FishExp", colnames(ifood))
colnames(ifood) <- gsub("MntSweetProducts", "SweetExp", colnames(ifood))
colnames(ifood) <- gsub("MntGoldProds", "GoldExp", colnames(ifood))
colnames(ifood) <- gsub("Marital_Status", "MaritalSts", colnames(ifood))
colnames(ifood) <- gsub("NumCatalogPurchases", "CatalogPurc", colnames(ifood))
colnames(ifood) <- gsub("AcceptedCmp1", "AccCmp1", colnames(ifood))
colnames(ifood) <- gsub("AcceptedCmp2", "AccCmp2", colnames(ifood))
colnames(ifood) <- gsub("AcceptedCmp3", "AccCmp3", colnames(ifood))
colnames(ifood) <- gsub("AcceptedCmp4", "AccCmp4", colnames(ifood))
colnames(ifood) <- gsub("AcceptedCmp5", "AccCmp5", colnames(ifood))
```
### 4. Handle outliers
```{r}
ifood$Age <- ifelse(ifood$Age > 80, 80, ifood$Age)
```
### 5. Handle missing values
```{r}
ifood <- ifood[!ifood$MaritalSts %in% c("YOLO", "Absurd"),]
ifood$MaritalSts[ifood$MaritalSts == "Alone"] <- "Single"
```
### 6. Impute missing Income using KNN
```{r}
ifood$Income <- ifelse(ifood$Income < 12500, NA, ifood$Income)

num_vars <- sapply(ifood, is.numeric)
complete_vars <- colnames(ifood)[num_vars]
missing_threshold <- 0.2 * nrow(ifood)
complete_vars <- complete_vars[colSums(is.na(ifood[, complete_vars])) < missing_threshold]
aux <- ifood[, complete_vars]

var <- "Income"
aux1 <- aux[!is.na(ifood[[var]]), , drop = FALSE]
aux2 <- aux[is.na(ifood[[var]]), , drop = FALSE]

cols_na <- colnames(aux2)[colSums(is.na(aux2)) > 0]
if (length(cols_na) > 0) {
  aux1 <- aux1[, !(colnames(aux1) %in% cols_na), drop = FALSE]
  aux2 <- aux2[, !(colnames(aux2) %in% cols_na), drop = FALSE]
}

knn_impute <- knn(aux1, aux2, ifood[[var]][!is.na(ifood[[var]])], k = 1)
ifood[[var]][is.na(ifood[[var]])] <- as.numeric(as.character(knn_impute))
```
### 7. Correct calculation of `TotAccCmp`
```{r}
ifood$TotAccCmp <- ifood$AccCmp1 + ifood$AccCmp2 + ifood$AccCmp3 + ifood$AccCmp4 + ifood$AccCmp5
```
### 8. Remove duplicate records
```{r}
ifood <- ifood %>% arrange(desc(Response)) %>% distinct_at(vars(-Response), .keep_all = TRUE)
```
### 9. Create `TotalExp` before using it
```{r}
ifood$TotalExp <- rowSums(ifood[, c("WineExp", "FruitExp", "MeatExp", "FishExp", "SweetExp", "GoldExp")], na.rm = TRUE)
```
### 10. Save cleaned dataset
```{r}
write.csv(ifood, "ifood_cleaned.csv", row.names = FALSE)
```

---

# **Variable Creation**

## *Second-Generation*

### Total Purchases
```{r}
ifood$TotalPurchases <- ifood$DealsPurc + ifood$WebPurc + ifood$CatalogPurc + ifood$StorePurc
```
### Purchase Frequency
```{r}
ifood$PurchaseFrequency <- ifelse(ifood$CustDays > 0, ifood$TotalPurchases / (ifood$CustDays / 30), 0)
```
### Preferred Product Category
```{r}
product_categories <- c("WineExp", "FruitExp", "MeatExp", "FishExp", "SweetExp", "GoldExp")
max_index <- apply(ifood[ , product_categories], 1, which.max)
ifood$PreferredProductCategory <- product_categories[max_index]
ifood$PreferredProductCategory <- as.factor(ifood$PreferredProductCategory)
```
### Preferred Purchase Channel
```{r}
channels <- c("DealsPurc", "WebPurc", "CatalogPurc", "StorePurc")
max_ch_index <- apply(ifood[ , channels], 1, which.max)
ifood$PreferredChannel <- channels[max_ch_index]
ifood$PreferredChannel <- as.factor(ifood$PreferredChannel)
```
### Average Spend Per Purchase
```{r}
ifood$AvgSpendPerPurchase <- ifelse(ifood$TotalPurchases > 0, ifood$TotalExp / ifood$TotalPurchases, 0)
```
### HasChildren
```{r}
ifood$HasChildren <- ifelse(ifood$Kidhome + ifood$Teenhome > 0, 1, 0)
```
### IncomeSegment
```{r}
income_quantiles <- quantile(ifood$Income, probs = c(0.33, 0.66), na.rm = TRUE)
ifood$IncomeSegment <- cut(ifood$Income, breaks = c(-Inf, income_quantiles[1], income_quantiles[2], Inf),
                           labels = c("Low", "Medium", "High"))
```
### CustomerTenure
```{r}
ifood$CustomerTenure <- ifood$CustDays / 365
```
### CampaignAcceptanceRate
```{r}
ifood$CampaignAcceptanceRate <- ifelse(ifood$TotAccCmp > 0, ifood$TotAccCmp / 5, 0)
```

## *Third-Generation*

### **Third-Generation Feature 1: Customer Segmentation via Clustering**

### Prepare data for clustering: use Recency, TotalPurchases (frequency), and TotalExp (monetary)
```{r}
cluster_data <- ifood %>% select(Recency, TotalPurchases, TotalExp)
```
### Scale the data for clustering
```{r}
cluster_data_scaled <- scale(cluster_data)
```
### Perform k-means clustering with 3 clusters (as an example)
```{r}
set.seed(123)  # for reproducibility
k3 <- kmeans(cluster_data_scaled, centers = 3, nstart = 25)  # nstart for better convergence
```
### Add the cluster assignment as a new feature
```{r}
ifood$CustomerSegment <- as.factor(k3$cluster)
```
#### *(Customers are now labeled 1, 2, or 3 based on their cluster segment)*

---

### **Third-Generation Feature 2: Propensity Score via Logistic Regression**
### Fit a logistic regression model to predict campaign response (Response) using relevant features
```{r}
propensity_model <- glm(Response ~ Income + Recency + TotalExp + TotalPurchases + TotAccCmp + Age + MaritalSts,
                        data = ifood, family = binomial)
```
### Get predicted probabilities (propensity to respond)
```{r}
ifood$PropensityScore <- predict(propensity_model, ifood, type = "response")
```
#### *(PropensityScore is the model's predicted probability of Response=1 for each customer)*
### Quick summary of PropensityScore range
```{r}
summary(ifood$PropensityScore)
```
---
### **Third-Generation Feature 3: Engagement Index**

### Normalize components between 0 and 1

#### *Note: For Recency, a lower value means more recent (more engaged), so we invert it.*
```{r}
recency_norm   <- (max(ifood$Recency) - ifood$Recency) / max(ifood$Recency)          # invert recency
frequency_norm <- ifood$TotalPurchases / max(ifood$TotalPurchases)                   # purchases normalized
monetary_norm  <- ifood$TotalExp / max(ifood$TotalExp)                               # spending normalized
campaign_norm  <- (ifood$TotAccCmp + ifood$Response) / 6                             # campaign acceptance (out of 6 campaigns total including last response)
webvisit_norm  <- ifood$WebVisits / max(ifood$WebVisits)                             # web visits normalized
```

### Calculate engagement index as average of all five components, scaled to 0-100
```{r}
ifood$EngagementIndex <- (recency_norm + frequency_norm + monetary_norm + campaign_norm + webvisit_norm) / 5 * 100
```
### Preview EngagementIndex distribution
```{r}
summary(ifood$EngagementIndex)
```
### Save enriched dataset
```{r}
write.csv(ifood, "ifood_enriched.csv", row.names = FALSE)
```
